Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-02-16 code/processing-code/processingfile-v1.qmd
Part 2 Updates
I processed several variables that we selected to answer our research question, updated readme files, and deleted extra files that are not related to our project. I found variables that seem to answer our first question already, so I formulated a second question in the manuscript.qmd file.

---
title: "Processing Script"
author: "Taylor Glass and Arlyn Santiago"
date: "2024-02-09"
output: html_document
---

@@

```{r}
library(readr) # for loading csv files
library(readxl) # for loading excel files (codebook)
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(naniar) #to explore data missingness
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

@@

# load codebook
codebook <- lapply(sheets, function(sheet) {
  readxl::read_excel(codebook_location, sheet = sheet)
})
print(codebook)
```

@@

First, the `region` variable is a character variable, but it will be more useful later as a factor. We can easily mutate it from a character variable to factor variable.  
```{r}
# convert character variable to factor variable
rawdata2 <- rawdata2 %>% 
              mutate(region = as.factor(region))

@@

Second, there are 24 entries missing for the percentage of all unintended pregnancies by country, which is only 18.18% of the observations. After isolating the observations with this missing variable, we see that the majority of countries in this category are located in Asia. This does not make sense because the majority of countries in the raw dataset are located in Africa, which suggests that the data are not missing at random. 
```{r}
# find missing percentage for each variable
miss_var_summary(rawdata2)

@@

We can impute the missing data here because only 18.18% of the data is missing, so the mean value of the variable is fairly reliable (find a source for this?) After finding the summary statistics of this variable, the median is 0.4701 and the mean is 0.4679, which suggests that there are minimal outliers in the data.  
```{r}
summary(rawdata2$pct_upreg)
```

Third, we need to evaluate the 108 missing entries of the percentage of unintended pregnancies subregion. Since 81.81% of this variable is missing, it does not make sense to include it in the analyis because we do not have enough information to impute the missing data. 
```{r}
# isolate the percentage of all unintended pregnancies by subregion to discover trends in missingness
variable_with_missing <- "upreg_subreg"
missing_obs2 <- rawdata2[is.na(rawdata2[[variable_with_missing]]), ]
other_variables2 <- colnames(rawdata2)[!colnames(rawdata2) == variable_with_missing]
pct_subregmissing <- missing_obs2[, other_variables2]
print(pct_subregmissing)
```

@@

```{r}
processeddata <- d4
```

@@

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed_data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

>>>>

---
title: "Processing Script"
author: "Taylor Glass and Arlyn Santiago"
date: "2024-02-16"
output: html_document
---

@@

```{r}
library(readr) # for loading csv files
library(readxl) # for loading excel files (codebook)
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(naniar) #to explore data missingness
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(tidymodels) #to impute missing data
library(recipes) #to impute missing data
```

@@

# load codebook
codebook <- lapply(sheets, function(sheet) {
  readxl::read_excel(codebook_location, sheet = sheet)
})
print(codebook)
```
Note: should we include other covariates like marriage percentages, number of live births, number of induced abortions

@@

First, the `region` variable is a character variable, but it will be more useful later as a factor. We can easily mutate it from a character variable to factor variable. All the remaining variables are numeric as they should be. 
```{r}
# convert character variable to factor variable
rawdata2 <- rawdata2 %>% 
              mutate(region = as.factor(region))

@@

Second, there are 24 entries missing for the percentage of all unintended pregnancies by country `pct_upreg`, which accounts for 18.18% of the observations. After isolating the observations with this missing variable, we see that the majority of countries in this category are located in Asia. This does not make sense because the majority of countries in the raw dataset are located in Africa, which suggests that the data are not missing at random. 
```{r}
# find missing percentage for each variable
miss_var_summary(rawdata2)

@@

We should impute the missing data here to avoid introducing bias because 18.18% of the data is missing, and this variable is the best outcome variable to answer the research question. After finding the summary statistics of this variable, the median is 0.4701 and the mean is 0.4679, which suggests that there are minimal outliers in the data.  
```{r}
# find mean to use for imputed data
summary(rawdata2$pct_upreg)

# create a recipe to impute data 
rec <- recipe(pct_upreg ~ ., data = rawdata2) %>%
  step_naomit(all_predictors(), -all_outcomes(), skip = TRUE) %>%
  step_impute_mean(all_predictors()) #which predictors would I use?

# prepare the raw data using the recipe
data_prep <- prep(rec, training = rawdata2)

# impute the data 
data_imputed <- bake(data_prep, new_data = NULL) #all pct_upreg are NaN values

# identify rows with missing values in the original dataset
missing_rows <- which(is.na(rawdata2$pct_upreg))

# replace missing values in the original dataset with imputed values
rawdata2$pct_upreg[missing_rows] <- data_imputed$pct_upreg[missing_rows]

```
Once the missing values have been imputed, we should further clean the `pct_upreg` variable to have a typical percentage format. 
```{r}
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg = pct_upreg * 100)
```

We also need to evaluate the 108 missing entries of the percentage of unintended pregnancies by subregion `upreg_subreg`. Since 81.81% of this variable is missing, it does not make sense to include it in the analysis because we do not have enough information to impute the missing data. We will remove this variable for the sake of this analysis.
```{r}
# isolate the percentage of all unintended pregnancies by subregion to discover trends in missingness
variable_with_missing <- "upreg_subreg"
missing_obs2 <- rawdata2[is.na(rawdata2[[variable_with_missing]]), ]
other_variables2 <- colnames(rawdata2)[!colnames(rawdata2) == variable_with_missing]
pct_subregmissing <- missing_obs2[, other_variables2]
print(pct_subregmissing)

# remove unintended pregnancies by subregion variable due to missingness
rawdata2 <- rawdata2 %>% 
              select(-upreg_subreg)
```

Each of the variables about birth control methods, `mod_methods3549`, `ster_3549`, `larc_3549`, `sarc_3549`, `tradmethods_3549`, `nomethod_3549`, are not very intuitive, so they should be renamed. Since we are only studying one age group, 35 to 49 years, it is unnecessary for those numbers to be included in each variable name. It also makes sense to rename `pct_upreg` to be more descriptive since we already know it is percentage format. 
```{r}
# rename predictor variables
rawdata2 <- rawdata2 %>% rename(modern = modmethods_3549,
                                sterile = ster_3549, 
                                longactrev = larc_3549,
                                shortact = sarc_3549,
                                tradition = tradmethods_3549,
                                nomethod = nomethod_3549,
                                unint_preg = pct_upreg)
# check if renaming worked
head(rawdata2)
```
To answer our research question about how unintended pregnancies are related to birth control methods, it makes sense to have a variable with the total number of women using birth control in each country. This will service as the main predictor variable, and each type of birth control can serve as a covariate. 
```{r}
# create new variable with total number on birth control of any form
rawdata2 <- rawdata2 %>% 
              mutate(totalbc = modern + sterile + longactrev + shortact + tradition + nomethod)

# check new variable is created 
head(rawdata2)
```

@@

```{r}
processeddata <- rawdata2
```

@@

Finally, we will save the clean data as RDS file to preserve coding like factors, characters, numeric, etc.

```{r}
save_data_location <- here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

### 2024-02-16 products/manuscript/Manuscript.qmd
Testing to see if I can push changes.
Added an extra backup question.
References
Part 2 Updates
I processed several variables that we selected to answer our research question, updated readme files, and deleted extra files that are not related to our project. I found variables that seem to answer our first question already, so I formulated a second question in the manuscript.qmd file.

Backup question: How do the rates of safe abortions fluctuate with available methods for safe abortions?

@@

>>>>

2) How does marriage status affect percentage of unintended pregnancies across different types of birth control?
  -outcome variables:
      unintended pregnancies, total number (pct_upreg)
      unintended pregnancies, using larc and sterilization (pct_upreg_larcster)
      unintended pregnancies, using short term reversible (pct_upreg_sarc)
  -predictor variables:
      Number wanting to avoid pregnancy, all married women (inneed_married)
      Number using sterilization, all married women	(ster_married) **combine these two in processing
      Number using long-acting reversible methods, all married women	(larc_married)
      Number using short-acting methods, all married women	(sarc_married)
      Number wanting to avoid pregnancy, formerly married women 	(inneed_formerlymar)
      Number using sterilization, formerly married women	(ster_formerlymar)
      Number using long-acting reversible methods, formerly married women	(larc_formerlymar)
      Number using short-acting methods, formerly married women	(sarc_formerlymar)
      Number wanting to avoid pregnancy, never married women 	(inneed_nevermar)
      Number using all modern methods, never married women	(modmethods_nevermar)
      Number using sterilization, never married women	(ster_nevermar)
      Number using long-acting reversible methods, never married women	(larc_nevermar)
      Number using short-acting methods, never married women	(sarc_nevermar)

@@

https://pubmed.ncbi.nlm.nih.gov/32981858/
https://pubmed.ncbi.nlm.nih.gov/30687521/ #This study is set in Senegal 
https://www.sciencedirect.com/science/article/pii/S001078241730478X
